---
title: "ORF Filtering Analysis — `r params$project$label`"
author: "Ribo-seq Pipeline"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
params:
  config: "config_template.yaml"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(data.table)
  library(scales)
  library(yaml)
  library(R6)
})

# Look for _functions.R relative to this qmd or in the pipeline root
if (file.exists("_functions.R")) {
  source("_functions.R")
} else if (file.exists("../../scripts/post_analysis/_functions.R")) {
  source("../../scripts/post_analysis/_functions.R")
} else {
  stop("Cannot find _functions.R")
}

cfg  <- load_config(params$config)
data <- load_pipeline_data(cfg)
df   <- data$merged
df   <- apply_expression_filters(df, cfg, data$expression)
```

# 1. Pipeline Overview

## 1.1 ORF Prediction Tools (Per-Tool Raw)

```{r raw-tools, echo=FALSE}
data.frame(
  Tool = c("Ribo-TISH", "ORFquant", "Ribotricer", "PRICE", "RiboCode"),
  Raw_ORFs = c("3,137,951", "1,342,002", "422,249", "359,166", "218,518"),
  Samples = c("23", "23", "23", "23", "23")
) |> knitr::kable(format="html", caption="Raw ORF predictions per tool (23 samples each)")
```

## 1.2 Unification

```
5,479,886 raw → exact-match dedup → 654,837 → frame-aware merge → 394,053 final
```

| Stage | ORFs | Reduction |
|-------|-----:|:---:|
| Raw (5 tools × 23 samples) | 5,479,886 | — |
| Exact-match merge | 654,837 | -88.0% |
| Frame-aware merge | **394,053** | -39.8% |

## 1.3 Classification (GENCODE)

| Biotype | Count | % |
|---------|------:|:---:|
| CDS | 121,069 | 66.0 |
| intORF | 27,355 | 14.9 |
| dORF | 19,691 | 10.7 |
| doORF | 10,195 | 5.6 |
| uORF | 3,117 | 1.7 |
| uoORF | 1,315 | 0.7 |
| lncRNA | 551 | 0.3 |
| **Total classified** | **183,293** | 46.5% of unified |

# 2. Data Sources & Methodology

## 1.1 Input Data

```{r overview}
data.frame(
  Source = c("Metadata", "Confidence (OCS)", "Expression (reads + pN)", "GENCODE"),
  N_ORFs = c(nrow(data$metadata), nrow(data$confidence),
             nrow(data$expression), nrow(data$gencode))
) |>
  knitr::kable(format = "html", caption = "Input data sources")

total_meta <- nrow(data$metadata)
with_gencode <- sum(!is.na(df$orf_biotype))
with_ocs     <- sum(!is.na(df$tier))
with_expr    <- sum(!is.na(df$total_reads_expr))

data.frame(
  Metric = c("Total unified ORFs", "With GENCODE classification",
             "With OCS score", "With expression data"),
  Count = c(total_meta, with_gencode, with_ocs, with_expr),
  Pct   = round(100 * c(total_meta, with_gencode, with_ocs, with_expr) / total_meta, 1)
) |>
  knitr::kable(format = "html", caption = "Cross-source match rates")
```

## 1.2 ORF Confidence Score (OCS) Methodology

The ORF Confidence Score (OCS) is a composite metric computed from four
independent sub-scores, each normalized to [0, 1]:

| Component | Source | Description |
|-----------|--------|-------------|
| **S_translation** | Tool-specific significance | Normalized p-value / score from the detecting tool (e.g. Ribo-TISH p-value, Ribotricer phase score). Aggregated across detecting tools by max. |
| **S_agreement** | Cross-tool overlap | Number of tools detecting the ORF, normalized by total tools. Multi-tool consensus increases confidence. |
| **S_coverage** | Expression breadth | Fraction of samples with reads > 0. Captures reproducibility across biological replicates. |
| **S_periodicity** | Read-level 3-nt periodicity | Aggregate periodicity score from RiboseQC read-length analysis. High periodicity indicates genuine ribosome footprints. |
| **S_readlevel** | Sample-wide QC modifier | Penalty factor when sample-level QC flags are present (e.g. low alignment rate, poor periodicity). |

**OCS Formula:**
$$OCS = w_t \cdot S_{translation} + w_a \cdot S_{agreement} + w_c \cdot S_{coverage} + w_p \cdot S_{periodicity} - S_{readlevel}$$

Default OCS weights: translation=0.35, agreement=0.25, coverage=0.25, periodicity=0.15.

**Tier thresholds:**
- **High**: OCS ≥ 0.8 — strong evidence, multiple tools, reproducible
- **Medium**: OCS ≥ 0.5 — moderate evidence
- **Low**: OCS ≥ 0.2 — weak but detectable signal
- **Uncertain**: OCS < 0.2 — likely noise

## 1.3 Expression-Based P-site Quality Proxies

Rather than rescanning bedgraph files (369K ORFs × 23 samples × 2 strands
≈ 109 CPU hours), we use per-sample `reads` and `pN` columns from
`unified_orfs_expression_summary.tsv` as computationally efficient proxies
for P-site quality:

| Original Criterion | Proxy | Rationale |
|-------------------|-------|-----------|
| `p_site_GSE > 9` | `{sample}_reads > 9` | Total reads overlapping the ORF correlates strongly with P-site count |
| `p_site_pct > 0.5` | `{sample}_pN > 0.5` | pN (periodicity score) is computed from P-site frame preference — high pN implies high P-site fraction |
| `p_site_pos > 2` | *dropped* | No direct proxy in expression summary; position-offset information is lost, but the first two criteria capture most of the signal |

**Cross-sample aggregation:** A per-sample criterion is "passed" if ANY
sample meets the threshold. More stringent strategies (≥2, ≥3 samples)
are tested in the filter comparison matrix below.

# 2. ORF Biotype Distribution

```{r biotype-dist}
biotype_counts <- df |>
  filter(!is.na(orf_biotype)) |>
  count(orf_biotype, sort = TRUE) |>
  mutate(orf_biotype = factor(orf_biotype, levels = rev(unique(orf_biotype))))

ggplot(biotype_counts, aes(n, orf_biotype, fill = orf_biotype == "CDS")) +
  geom_col() +
  scale_x_log10(labels = comma) +
  scale_fill_manual(values = c("TRUE" = "grey60", "FALSE" = "#E64B35"), guide = "none") +
  labs(title = "ORF biotype distribution (GENCODE classification)",
       subtitle = sprintf("Non-CDS: %s ORFs across %d biotypes",
         comma(sum(biotype_counts$n[biotype_counts$orf_biotype != "CDS"])),
         nrow(biotype_counts) - 1),
       x = "Number of ORFs (log10)", y = NULL) +
  theme_minimal(base_size = 13)
```

# 3. OCS Tier Distribution

```{r ocs-tier}
df |>
  filter(!is.na(tier)) |>
  count(tier) |>
  mutate(tier = factor(tier, levels = c("High", "Medium", "Low", "Uncertain")),
         pct = n / sum(n)) |>
  ggplot(aes(tier, n, fill = tier)) +
  geom_col() +
  geom_text(aes(label = sprintf("%s\n(%.1f%%)", comma(n), 100 * pct)), vjust = -0.2) +
  scale_fill_manual(values = c(High = "#00A087", Medium = "#7E6148",
                                Low = "#E64B35", Uncertain = "grey70"), guide = "none") +
  labs(title = "OCS tier distribution", x = NULL, y = "Count") +
  theme_minimal(base_size = 13)
```

# 4. Expression Distributions

```{r expr-dist, fig.height=10, fig.width=12}
read_cols <- grep("_reads$", names(data$expression), value = TRUE)
pN_cols   <- grep("_pN$", names(data$expression), value = TRUE)

expr_agg <- data$expression |>
  reframe(
    max_reads = do.call(pmax, c(pick(all_of(read_cols)), na.rm = TRUE)),
    max_pN    = do.call(pmax, c(pick(all_of(pN_cols)), na.rm = TRUE)),
    n_samples = rowSums(pick(all_of(read_cols)) > 0, na.rm = TRUE)
  )

# Reads histogram
p1 <- ggplot(expr_agg, aes(max_reads)) +
  geom_histogram(bins = 100, fill = "#4682B4", alpha = 0.8) +
  geom_vline(xintercept = cfg$filtering$p_site_gse_min, color = "#E64B35",
             linewidth = 1, linetype = "dashed") +
  scale_x_log10(labels = comma) +
  labs(title = "Max reads per ORF (proxy for P-site GSE count)",
       subtitle = sprintf("Red line = threshold (reads > %d)", cfg$filtering$p_site_gse_min),
       x = "max reads across samples (log10)", y = "Count") +
  theme_minimal(base_size = 12)

# pN histogram
p2 <- ggplot(expr_agg, aes(max_pN)) +
  geom_histogram(bins = 100, fill = "#7E6148", alpha = 0.8) +
  geom_vline(xintercept = cfg$filtering$p_site_pct_min, color = "#E64B35",
             linewidth = 1, linetype = "dashed") +
  labs(title = "Max pN per ORF (proxy for P-site fraction)",
       subtitle = sprintf("Red line = threshold (pN > %.1f)", cfg$filtering$p_site_pct_min),
       x = "max pN across samples", y = "Count") +
  theme_minimal(base_size = 12)

# Sample coverage
p3 <- ggplot(expr_agg, aes(n_samples)) +
  geom_histogram(binwidth = 1, fill = "#00A087", alpha = 0.8) +
  labs(title = "Samples with reads per ORF",
       x = "N samples with reads > 0", y = "Count") +
  theme_minimal(base_size = 12)

print(p1); print(p2); print(p3)
```

# 5. Filter Strategy Comparison

```{r filter-matrix}
filter_matrix <- build_filter_matrix(df)

filter_matrix |>
  select(filter_label, n_retained, pct_total, n_OCS_High, n_OCS_Medium,
         n_nonCDS, n_multi_tool, median_pN) |>
  knitr::kable(format = "html", digits = 1,
    col.names = c("Strategy", "Retained", "% Total", "High OCS",
                  "Medium OCS", "Non-CDS", "≥2 tools", "Median pN"),
    caption = "Filter strategy comparison matrix")

# Visual
filter_matrix |>
  mutate(filter_label = factor(filter_label, levels = filter_label[order(n_retained)])) |>
  ggplot(aes(n_retained, n_OCS_High)) +
  geom_point(aes(size = n_nonCDS, color = pct_total), alpha = 0.8) +
  geom_text(aes(label = filter_label), size = 3, hjust = -0.1, vjust = 0.5) +
  scale_x_log10(labels = comma) +
  scale_color_viridis_c() +
  labs(title = "Filter strategies: retention vs OCS High overlap",
       x = "ORFs retained (log10)", y = "High OCS retained",
       size = "Non-CDS", color = "% Total") +
  theme_minimal(base_size = 12)
```

# 6. Orthogonality: Expression Filters × OCS

```{r orthog}
# Non-CDS cross-tabulation
ct <- df |>
  filter(!is.na(orf_biotype), orf_biotype != "CDS") |>
  mutate(
    filter_status = case_when(
      pass_triple_3 ~ "Triple (≥3 samples)",
      pass_triple_2 ~ "Triple (≥2 samples)",
      pass_triple_any ~ "Triple (any sample)",
      TRUE ~ "Not passed"
    ),
    ocs_group = case_when(
      tier == "High"               ~ "OCS High",
      tier %in% c("High","Medium") ~ "OCS Medium+",
      TRUE                          ~ "OCS Low/Uncertain"
    )
  ) |>
  count(filter_status, ocs_group) |>
  tidyr::pivot_wider(names_from = ocs_group, values_from = n, values_fill = 0)

knitr::kable(ct, format = "html",
  caption = "Non-CDS ORFs: expression filter × OCS tier cross-tabulation")

# Tool agreement breakdown
ct2 <- df |>
  filter(!is.na(orf_biotype), orf_biotype != "CDS") |>
  mutate(pass = ifelse(pass_triple_2, "Pass filter", "Fail filter")) |>
  count(pass, n_detecting) |>
  tidyr::pivot_wider(names_from = n_detecting, values_from = n, values_fill = 0,
                     names_prefix = "Ntools=")

knitr::kable(ct2, format = "html",
  caption = "Non-CDS ORFs: expression filter × number of detecting tools")
```

# 7. Correlation Between Scoring Dimensions

```{r cor-heatmap, fig.height=6, fig.width=9}
cor_data <- df |>
  filter(!is.na(orf_biotype), orf_biotype != "CDS") |>
  select(pN, total_reads_expr, max_pN, n_samples,
         s_translation, s_agreement, s_coverage, s_periodicity,
         n_detecting, n_pass_all) |>
  rename(
    `pN (metadata)` = pN,    `Total reads` = total_reads_expr,
    `Max pN` = max_pN,       `N samples` = n_samples,
    S_translation = s_translation, S_agreement = s_agreement,
    S_coverage = s_coverage, S_periodicity = s_periodicity,
    `N tools` = n_detecting, `N pass filter` = n_pass_all
  )

cor_matrix <- cor(cor_data, use = "pairwise.complete.obs")

cor_matrix |>
  as.data.frame() |>
  as.data.frame() |>
  mutate(var1 = rownames(cor_matrix)) |>
  select(var1, everything()) |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  ggplot(aes(var1, var2, fill = r)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  scale_fill_gradient2(low = "#3B4992", mid = "white", high = "#E64B35",
                       limits = c(-1, 1)) +
  labs(title = "Correlation: scoring dimensions (non-CDS ORFs)",
       subtitle = "Expression metrics (reads, pN) are largely orthogonal to OCS sub-scores",
       x = NULL, y = NULL, fill = "r") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 8. Final ORF Selection

**Filter:** non-CDS + reads>9 + pN>0.5 (≥2 samples) + length ≤50 amino acids

```{r final-selection}
final_orfs <- df |>
  filter(!is.na(orf_biotype), orf_biotype != "CDS",
         pass_triple_2, length_aa <= 50)

message(sprintf("Final ggRibo candidates: %s ORFs", comma(nrow(final_orfs))))

final_orfs |>
  count(orf_biotype, sort = TRUE) |>
  knitr::kable(format = "html", col.names = c("Biotype", "Count"),
               caption = "Per-biotype ORF counts passing all filters")

# Write full list (no sampling)
out_tsv <- file.path(cfg$input$output_dir, "final_orfs_for_ggribo.tsv")
final_orfs |>
  select(orf_id, chrom, start, end, strand, orf_biotype,
         total_reads_expr, max_pN, tier, n_detecting) |>
  fwrite(out_tsv, sep = "\t")

message(sprintf("Wrote %s ORFs → %s", comma(nrow(final_orfs)), out_tsv))
```

# 9. OCS Distribution for Final ORFs

```{r ocs-by-biotype, fig.height=8, fig.width=10}
# OCS score distribution for final selected ORFs
hist(final_orfs$ocs, breaks = 60, col = "#4682B4", border = "white",
     main = "OCS Distribution of Final Selected ORFs",
     xlab = "ORF Confidence Score (OCS)", ylab = "Count")
abline(v = c(0.2, 0.5, 0.8), lty = 2, col = "#E64B35", lwd = 2)
text(c(0.2, 0.5, 0.8), par("usr")[4] * 0.95,
     labels = c("Low", "Medium", "High"), pos = 3, col = "#E64B35", cex = 0.9)

# OCS boxplot by biotype
boxplot(ocs ~ orf_biotype, data = final_orfs,
        col = RColorBrewer::brewer.pal(6, "Set2"),
        main = "OCS Distribution by ORF Biotype",
        xlab = NULL, ylab = "ORF Confidence Score (OCS)",
        las = 2, cex.axis = 0.9)
abline(h = c(0.2, 0.5, 0.8), lty = 2, col = "grey60")
```

```{r ocs-tier-biotype, fig.height=5, fig.width=10}
# Tier composition by biotype
tt <- as.data.frame(table(final_orfs$orf_biotype, final_orfs$tier))
names(tt) <- c("biotype", "tier", "n")
tt$biotype <- factor(tt$biotype, levels = c("uORF","dORF","doORF","intORF","uoORF","lncRNA"))
tt$tier    <- factor(tt$tier, levels = c("High","Medium","Low","Uncertain"))

ggplot(tt, aes(biotype, n, fill = tier)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = c(High = "#00A087", Medium = "#7E6148",
                                Low = "#E64B35", Uncertain = "grey70"),
                    name = "OCS Tier") +
  labs(title = "OCS tier composition by ORF biotype",
       subtitle = "Final selected ORFs (non-CDS, reads>9+pN>0.5(≥2), ≤50aa)",
       x = NULL, y = "Proportion") +
  theme_minimal(base_size = 12)
```

# 10. ORF Length Distribution

```{r length-dist, fig.height=8, fig.width=12}
# Pre-filter: all classified ORFs
p1 <- df |>
  filter(!is.na(orf_biotype)) |>
  ggplot(aes(length_aa, fill = orf_biotype == "CDS")) +
  geom_histogram(bins = 80, alpha = 0.7) +
  geom_vline(xintercept = 50, color = "#E64B35", linewidth = 1, linetype = "dashed") +
  scale_fill_manual(values = c("TRUE" = "grey60", "FALSE" = "#4682B4"),
                    labels = c("TRUE" = "CDS", "FALSE" = "Non-CDS"), name = NULL) +
  scale_x_log10(labels = comma) +
  labs(title = "ORF length distribution — all classified ORFs (pre-filter)",
       subtitle = sprintf("Red line: 50aa cutoff. Median = %s aa, %s ORFs ≤ 50aa",
         comma(median(df$length_aa, na.rm = TRUE)),
         comma(sum(df$length_aa <= 50, na.rm = TRUE))),
       x = "ORF length (amino acids, log10)", y = "Count") +
  theme_minimal(base_size = 12)

# Post-filter: final selected ORFs only
p2 <- final_orfs |>
  ggplot(aes(length_aa, fill = orf_biotype)) +
  geom_histogram(bins = 50, alpha = 0.8) +
  facet_wrap(~ orf_biotype, scales = "free_y", ncol = 3) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(title = "ORF length distribution — final selected (post-filter)",
       subtitle = sprintf("Non-CDS, reads>9+pN>0.5(≥2), ≤50aa. %s ORFs",
         comma(nrow(final_orfs))),
       x = "ORF length (amino acids)", y = "Count") +
  theme_minimal(base_size = 12)

# Pre vs post comparison by biotype
pre_len <- df |>
  filter(!is.na(orf_biotype), orf_biotype != "CDS") |>
  select(orf_biotype, length_aa) |>
  mutate(stage = "Pre-filter")
post_len <- final_orfs |>
  select(orf_biotype, length_aa) |>
  mutate(stage = "Post-filter")
biotype_len <- dplyr::bind_rows(pre_len, post_len)

p3 <- biotype_len |>
  ggplot(aes(stage, length_aa, fill = stage)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  facet_wrap(~ orf_biotype, ncol = 3) +
  scale_fill_manual(values = c("Pre-filter" = "grey70", "Post-filter" = "#00A087"),
                    guide = "none") +
  labs(title = "Length distribution: pre vs post filter by biotype",
       x = NULL, y = "ORF length (amino acids)") +
  theme_minimal(base_size = 12)

print(p1)
print(p2)
print(p3)
```

# 11. Summary

```{r summary}
data.frame(
  Stage = c("Total unified ORFs", "With GENCODE", "Non-CDS",
            "reads>9+pN>0.5 (any)", "reads>9+pN>0.5 (≥2)", "reads>9+pN>0.5 (≥3)",
            "Final selected (≤50aa)"),
  Count = c(nrow(data$metadata), sum(!is.na(df$orf_biotype)),
            sum(!is.na(df$orf_biotype) & df$orf_biotype != "CDS"),
            sum(df$pass_triple_any, na.rm = TRUE),
            sum(df$pass_triple_2, na.rm = TRUE),
            sum(df$pass_triple_3, na.rm = TRUE),
            nrow(final_orfs))
) |>
  knitr::kable(format = "html",
    caption = sprintf("Filtering cascade — %s", cfg$project$label))
```

## Recommended Filter

**Non-CDS + reads>9 + pN>0.5 (≥2 samples) + length ≤50 amino acids.**

This combines signal strength (reads), periodicity quality (pN),
reproducibility (≥2 samples), and sORF length constraint (≤50aa).
The filter is largely orthogonal to the OCS scoring system
(correlation r < 0.3), providing independent validation.

**Final selected:** `{r} comma(nrow(final_orfs))` ORFs across 6 biotypes.
