---
title: "ORF Filtering Analysis — `r params$project$label`"
author: "Ribo-seq Pipeline"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
params:
  config: "config_template.yaml"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(data.table)
  library(scales)
  library(yaml)
  library(R6)
})
# Optional: for richer plots
has_patchwork <- requireNamespace("patchwork", quietly = TRUE)
has_ggrepel   <- requireNamespace("ggrepel", quietly = TRUE)
has_UpSetR    <- requireNamespace("UpSetR", quietly = TRUE)

source(here::here("scripts/post_analysis/_functions.R"))

cfg <- load_config(params$config)
data <- load_pipeline_data(cfg)
df   <- data$merged

# Apply expression-based quality filters (no bedgraph rescanning)
df <- apply_expression_filters(df, cfg, data$expression)
```

# 1. Data Overview

```{r overview}
tibble(
  Source = c("Metadata", "Confidence (OCS)", "Expression (reads + pN)", "GENCODE"),
  N_ORFs = c(
    nrow(data$metadata),
    nrow(data$confidence),
    nrow(data$expression),
    nrow(data$gencode)
  )
) |>
  mutate(Source = fct_inorder(Source)) |>
  knitr::kable(format = "html", caption = "Input data sources")

# Merge rate
total_meta <- nrow(data$metadata)
with_gencode <- sum(!is.na(df$orf_biotype))
with_ocs     <- sum(!is.na(df$tier))
with_expr    <- sum(!is.na(df$total_reads_expr))

tibble(
  Metric = c("Total unified ORFs", "With GENCODE classification",
             "With OCS score", "With expression data"),
  Count = c(total_meta, with_gencode, with_ocs, with_expr),
  Pct = 100 * c(total_meta, with_gencode, with_ocs, with_expr) / total_meta
) |>
  knitr::kable(format = "html", digits = 1,
               caption = "Cross-source match rates")
```

# 2. ORF Biotype Distribution

```{r biotype-dist}
biotype_counts <- df |>
  filter(!is.na(orf_biotype)) |>
  count(orf_biotype, sort = TRUE) |>
  mutate(orf_biotype = fct_reorder(orf_biotype, n))

ggplot(biotype_counts, aes(n, orf_biotype, fill = orf_biotype == "CDS")) +
  geom_col() +
  scale_x_log10(labels = scales::comma) +
  scale_fill_manual(values = c("TRUE" = "grey60", "FALSE" = "#E64B35"), guide = "none") +
  labs(title = "ORF biotype distribution (GENCODE)",
       x = "Number of ORFs (log10)", y = NULL) +
  theme_minimal(base_size = 13)
```

**Non-CDS ORFs:** `{r} scales::comma(sum(biotype_counts$n[biotype_counts$orf_biotype != "CDS"]))` across {nrow(biotype_counts) - 1} biotypes.

# 3. OCS Tier Distribution

```{r ocs-tier}
df |>
  filter(!is.na(tier)) |>
  count(tier) |>
  mutate(tier = fct_relevel(tier, "High", "Medium", "Low", "Uncertain"),
         pct = n / sum(n)) |>
  ggplot(aes(tier, n, fill = tier)) +
  geom_col() +
  geom_text(aes(label = sprintf("%s\n(%.1f%%)", scales::comma(n), 100 * pct)),
            vjust = -0.2) +
  scale_fill_manual(values = c(High = "#00A087", Medium = "#7E6148",
                                Low = "#E64B35", Uncertain = "grey70"),
                    guide = "none") +
  labs(title = "OCS tier distribution", x = NULL, y = "Count") +
  theme_minimal(base_size = 13)
```

# 4. Per-Sample Expression Distributions (reads × pN)

```{r expr-dist, fig.height=10, fig.width=12}
read_cols <- grep("_reads$", names(data$expression), value = TRUE)
pN_cols   <- grep("_pN$", names(data$expression), value = TRUE)

# Per-ORF max reads and max pN across samples
expr_agg <- data$expression |>
  mutate(
    max_reads = apply(pick(all_of(read_cols)), 1, max, na.rm = TRUE),
    max_pN    = apply(pick(all_of(pN_cols)), 1, max, na.rm = TRUE),
    n_samples = apply(pick(all_of(read_cols)), 1, function(x) sum(x > 0, na.rm = TRUE))
  )

thresholds <- tibble(
  metric = c("max_reads", "max_pN"),
  label  = c("Max reads per ORF", "Max pN per ORF"),
  thresh = c(cfg$filtering$p_site_gse_min, cfg$filtering$p_site_pct_min)
)

p1 <- bind_rows(
  expr_agg |> select(value = max_reads) |> mutate(metric = "max_reads"),
  expr_agg |> select(value = max_pN) |> mutate(metric = "max_pN")
) |>
  left_join(thresholds, by = "metric") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 100, fill = "#4682B4", alpha = 0.8) +
  geom_vline(aes(xintercept = thresh), color = "#E64B35",
             linewidth = 1, linetype = "dashed") +
  facet_wrap(~ label, scales = "free", ncol = 1) +
  scale_x_log10(labels = scales::comma) +
  labs(title = "Per-ORF expression distributions (proxy for P-site purity)",
       subtitle = sprintf("Red line = filter threshold (reads>%d, pN>%.1f)",
                cfg$filtering$p_site_gse_min, cfg$filtering$p_site_pct_min),
       x = NULL, y = "Count") +
  theme_minimal(base_size = 12)

# Sample coverage histogram
p2 <- expr_agg |>
  ggplot(aes(n_samples)) +
  geom_histogram(binwidth = 1, fill = "#7E6148", alpha = 0.8) +
  labs(title = "Number of samples with reads per ORF",
       x = "N samples with reads > 0", y = "Count") +
  theme_minimal(base_size = 12)

print(p1 / p2 + plot_layout(heights = c(2, 1)))
```

# 5. Filter Strategy Comparison

```{r filter-matrix}
filter_matrix <- build_filter_matrix(df)

filter_matrix |>
  select(filter_label, n_retained, pct_total, n_OCS_High, n_OCS_Medium,
         n_nonCDS, n_multi_tool, median_pN) |>
  knitr::kable(
    format = "html",
    digits = 1,
    col.names = c("Strategy", "Retained", "% Total", "High OCS",
                  "Medium OCS", "Non-CDS", "≥2 tools", "Median pN"),
    caption = "Filter strategy comparison matrix"
  )

# Visual: retained vs OCS overlap
filter_matrix |>
  mutate(filter_label = fct_reorder(filter_label, n_retained)) |>
  ggplot(aes(n_retained, n_OCS_High, label = filter_label)) +
  geom_point(aes(size = n_nonCDS, color = pct_total)) +
  geom_text_repel(size = 3, max.overlaps = 15) +
  scale_x_log10(labels = scales::comma) +
  scale_color_viridis_c() +
  labs(title = "Filter strategies: retention vs OCS High overlap",
       x = "ORFs retained (log10)", y = "High OCS retained",
       size = "Non-CDS", color = "% Total") +
  theme_minimal(base_size = 12)
```

# 6. Orthogonality: Expression Filters × OCS

```{r upset, fig.height=8, fig.width=12}
upset_data <- df |>
  filter(!is.na(orf_biotype), orf_biotype != "CDS") |>
  mutate(
    `reads>9 + pN>0.5 (any)` = as.integer(pass_triple_any),
    `reads>9 + pN>0.5 (≥2)` = as.integer(pass_triple_2),
    `reads>9 + pN>0.5 (≥3)` = as.integer(pass_triple_3),
    `OCS High`               = as.integer(tier == "High"),
    `OCS Medium+`            = as.integer(tier %in% c("High", "Medium")),
    `≥2 detecting tools`     = as.integer(n_detecting >= 2),
    `High expression`        = as.integer(total_reads_expr >= quantile(total_reads_expr, 0.9, na.rm = TRUE))
  ) |>
  select(orf_id, `reads>9 + pN>0.5 (any)`, `reads>9 + pN>0.5 (≥2)`,
         `reads>9 + pN>0.5 (≥3)`, `OCS High`, `OCS Medium+`,
         `≥2 detecting tools`, `High expression`) |>
  column_to_rownames("orf_id") |>
  as.data.frame()

n_up <- nrow(upset_data)
if (n_up > 5000) {
  upset_data <- upset_data[sample(n_up, 5000), ]
}

upset(upset_data,
      sets = c("reads>9 + pN>0.5 (≥2)", "OCS High", "≥2 detecting tools",
               "High expression"),
      order.by = "freq",
      main.bar.color = "#4682B4",
      sets.bar.color = "#7E6148",
      text.scale = 1.5)
```

# 7. Correlation Between Scoring Dimensions

```{r cor-heatmap, fig.height=6, fig.width=8}
cor_data <- df |>
  filter(!is.na(orf_biotype), orf_biotype != "CDS") |>
  select(
    pN, total_reads_expr, max_pN, mean_ratio, n_samples,
    s_translation, s_agreement, s_coverage, s_periodicity,
    n_detecting, n_pass_all
  ) |>
  rename(
    `pN (metadata)` = pN,
    `Total reads` = total_reads_expr,
    `Max pN` = max_pN,
    `Mean ratio` = mean_ratio,
    `N samples` = n_samples,
    `S_translation` = s_translation,
    `S_agreement` = s_agreement,
    `S_coverage` = s_coverage,
    `S_periodicity` = s_periodicity,
    `N detecting tools` = n_detecting,
    `N pass triple` = n_pass_all
  )

cor_matrix <- cor(cor_data, use = "pairwise.complete.obs")

cor_matrix |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  ggplot(aes(var1, var2, fill = r)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  scale_fill_gradient2(low = "#3B4992", mid = "white", high = "#E64B35",
                       limits = c(-1, 1)) +
  labs(title = "Correlation: scoring dimensions (non-CDS ORFs)",
       x = NULL, y = NULL, fill = "Pearson r") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 8. Final ORF Selection for ggRibo

```{r final-selection}
# Apply the recommended default filter: triple-filter (≥2 samples), non-CDS
final_orfs <- df |>
  filter(
    !is.na(orf_biotype),
    orf_biotype != "CDS",
    pass_triple_2
  )

n_final <- nrow(final_orfs)
message(sprintf("Final ggRibo candidates: %s ORFs", scales::comma(n_final)))

# Per-biotype counts
final_orfs |>
  count(orf_biotype, sort = TRUE) |>
  knitr::kable(format = "html", col.names = c("Biotype", "Count"),
               caption = "Per-biotype ORF counts for ggRibo")

# Write final ORF list
ggribo_cfg <- cfg$ggribo
final_orfs_sampled <- final_orfs |>
  slice_max(total_reads_expr, n = ggribo_cfg$max_per_biotype,
            by = orf_biotype, with_ties = FALSE)

out_tsv <- file.path(cfg$input$output_dir, "final_orfs_for_ggribo.tsv")
final_orfs_sampled |>
  select(orf_id, chrom, start, end, strand, orf_biotype,
         total_reads_expr, max_pN, tier, n_detecting) |>
  write_tsv(out_tsv)

message(sprintf("  Sampled %s ORFs (max %d per biotype) → %s",
        scales::comma(nrow(final_orfs_sampled)),
        ggribo_cfg$max_per_biotype, out_tsv))
```

# 9. Summary & Recommendations

```{r summary}
tibble(
  Stage = c("Total unified ORFs", "With GENCODE", "Non-CDS",
            "Triple-filter (any sample)", "Triple-filter (≥2 samples)",
            "Final ggRibo sampled"),
  Count = c(
    nrow(data$metadata),
    sum(!is.na(df$orf_biotype)),
    sum(!is.na(df$orf_biotype) & df$orf_biotype != "CDS"),
    sum(df$pass_triple_any, na.rm = TRUE),
    sum(df$pass_triple_2, na.rm = TRUE),
    nrow(final_orfs_sampled)
  )
) |>
  knitr::kable(format = "html",
    caption = sprintf("Filtering cascade — %s", cfg$project$label))
```

**Recommended default filter:** Triple-filter (GSE>9 + pct>0.5 + pos>2) in ≥2 samples, non-CDS only. This balances stringency with sample coverage and is orthogonal to the OCS scoring system.
