---
title: "ORF Preliminary Analysis — `r params$project$label`"
author: "Ribo-seq Pipeline"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
params:
  config: "config_template.yaml"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(data.table)
  library(scales)
  library(yaml)
  library(R6)
})

# Look for _functions.R relative to this qmd or in the pipeline root
if (file.exists("_functions.R")) {
  source("_functions.R")
} else if (file.exists("../../scripts/post_analysis/_functions.R")) {
  source("../../scripts/post_analysis/_functions.R")
} else {
  stop("Cannot find _functions.R")
}

cfg  <- load_config(params$config)
data <- load_pipeline_data(cfg)
df   <- data$merged
df   <- apply_prelim_filters(df, cfg, data$expression)
```

# 1. GENCODE Classification Overview

## 1.1 Category Breakdown

The GENCODE mapper classifies ORFs by intersecting their coordinates with
transcript exon chains. ORFs that don't overlap any annotated transcript
are labelled **"intergenic"** — they are inherently non-CDS since there is
no transcript CDS to overlap.

```{r gencode-cats, echo=FALSE}
# Biotype distribution including intergenic
biotype_all <- df |>
  count(orf_biotype, sort = TRUE) |>
  mutate(pct = 100 * n / sum(n),
         orf_biotype = factor(orf_biotype, levels = rev(unique(orf_biotype))))

biotype_all |>
  knitr::kable(format = "html", digits = 1,
    col.names = c("Biotype", "Count", "%"),
    caption = "Full ORF biotype distribution (including intergenic)")
```

```{r gencode-bar, fig.height=5, fig.width=10, echo=FALSE}
biotype_colors <- c(
  CDS        = "grey70",
  intORF     = "#E64B35",
  dORF       = "#00A087",
  doORF      = "#7E6148",
  uORF       = "#4682B4",
  uoORF      = "#F39B7F",
  lncRNA     = "#3C5488",
  intergenic = "#FFDC91"
)

ggplot(biotype_all, aes(n, orf_biotype, fill = orf_biotype)) +
  geom_col() +
  scale_x_log10(labels = comma) +
  scale_fill_manual(values = biotype_colors, guide = "none") +
  labs(title = "ORF biotype distribution (GENCODE + intergenic)",
       subtitle = sprintf("Classified: %s (%.1f%%) | Intergenic: %s (%.1f%%) | Non-CDS: %s",
         comma(sum(df$is_classified)),
         100 * mean(df$is_classified),
         comma(sum(df$orf_biotype == "intergenic")),
         100 * mean(df$orf_biotype == "intergenic"),
         comma(sum(df$orf_biotype != "CDS", na.rm = TRUE))),
       x = "Number of ORFs (log10)", y = NULL) +
  theme_minimal(base_size = 13)
```

```{r classified-vs-intergenic, echo=FALSE}
# Summary table
data.frame(
  Category = c("Classified (GENCODE)", "Intergenic (unclassified)", "Total unified"),
  Count = c(sum(df$is_classified), sum(df$orf_biotype == "intergenic"), nrow(df)),
  Pct = c(100 * mean(df$is_classified), 100 * mean(df$orf_biotype == "intergenic"), 100)
) |>
  knitr::kable(format = "html", digits = 1,
    caption = "Classification coverage")
```

## 1.2 ORF Prediction Pipeline

```
5,479,886 raw → exact-match dedup → 654,837 → frame-aware merge → 394,053 unified
  → 183,293 GENCODE classified (46.5%) + 210,760 intergenic (53.5%)
```

| Stage | ORFs | Reduction |
|-------|-----:|:---:|
| Raw (5 tools × 23 samples) | 5,479,886 | — |
| Exact-match merge | 654,837 | -88.0% |
| Frame-aware merge | **394,053** | -39.8% |
| GENCODE classified | 183,293 | 46.5% of unified |
| Intergenic (unclassified) | 210,760 | 53.5% of unified |

# 2. Data Sources & Methodology

## 2.1 Input Data

```{r overview, echo=FALSE}
data.frame(
  Source = c("Metadata", "Confidence (OCS)", "Expression (reads + pN)", "GENCODE"),
  N_ORFs = c(nrow(data$metadata), nrow(data$confidence),
             nrow(data$expression), nrow(data$gencode))
) |>
  knitr::kable(format = "html", caption = "Input data sources")

total_meta <- nrow(data$metadata)
with_gencode <- sum(df$is_classified)
with_ocs     <- sum(!is.na(df$tier))
with_expr    <- sum(!is.na(df$total_reads_expr))

data.frame(
  Metric = c("Total unified ORFs", "With GENCODE classification",
             "With OCS score", "With expression data"),
  Count = c(total_meta, with_gencode, with_ocs, with_expr),
  Pct   = round(100 * c(total_meta, with_gencode, with_ocs, with_expr) / total_meta, 1)
) |>
  knitr::kable(format = "html", caption = "Cross-source match rates")
```

## 2.2 ORF Confidence Score (OCS) Methodology

The ORF Confidence Score (OCS) is a composite metric computed from four
independent sub-scores, each normalized to [0, 1]:

| Component | Source | Description |
|-----------|--------|-------------|
| **S_translation** | Tool-specific significance | Normalized p-value / score from the detecting tool |
| **S_agreement** | Cross-tool overlap | Number of tools detecting the ORF |
| **S_coverage** | Expression breadth | Fraction of samples with reads > 0 |
| **S_periodicity** | Read-level 3-nt periodicity | Aggregate periodicity score from RiboseQC |
| **S_readlevel** | Sample-wide QC modifier | Penalty factor for low-quality samples |

**Tier thresholds:**
- **High**: OCS ≥ 0.8 — strong evidence, multiple tools, reproducible
- **Medium**: OCS ≥ 0.5 — moderate evidence
- **Low**: OCS ≥ 0.2 — weak but detectable signal
- **Uncertain**: OCS < 0.2 — likely noise

## 2.3 Two-Stage Filtering Strategy

This analysis implements a **two-stage filtering** pipeline:

### Stage 1: Preliminary (expression-based, this report)
Uses `{sample}_reads` and `{sample}_pN` from the expression summary as
computationally cheap filters to eliminate ORFs with insufficient signal
before running the expensive bedgraph scan:

| Filter | Column | Threshold | Rationale |
|--------|--------|:---------:|-----------|
| Signal presence | `{sample}_reads` | > `r cfg$filtering$prelim_reads_min` | Validated r=0.99 vs real `p_site_GSE` |
| Periodicity | `{sample}_pN` | > `r cfg$filtering$prelim_pN_min` | Frame periodicity quality (NOT P-site fraction proxy; r=-0.20 vs real `p_site_pct`) |
| Cross-sample | — | ≥`r cfg$filtering$prelim_cross_sample` sample | Samples from heterogeneous datasets — any sample sufficient |

### Stage 2: Real P-site (bedgraph-based, next step)
`compute_psite_purity.py` backtracks through actual RiboseQC bedgraph
files to compute exact `p_site_GSE`, `p_site_pct`, and `p_site_pos`
per ORF per sample — no proxies, no approximations. This runs on the
**reduced ORF set** from Stage 1, dramatically cutting runtime.

# 3. ORF Biotype Distribution

```{r biotype-dist, fig.height=6, fig.width=10, echo=FALSE}
ggplot(biotype_all, aes(n, orf_biotype, fill = orf_biotype)) +
  geom_col() +
  scale_x_log10(labels = comma) +
  scale_fill_manual(values = biotype_colors, guide = "none") +
  labs(title = "ORF biotype distribution",
       subtitle = sprintf("Non-CDS (incl. intergenic): %s ORFs",
         comma(sum(df$orf_biotype != "CDS", na.rm = TRUE))),
       x = "Number of ORFs (log10)", y = NULL) +
  theme_minimal(base_size = 13)
```

# 4. OCS Tier Distribution

```{r ocs-tier, echo=FALSE}
df |>
  filter(!is.na(tier)) |>
  count(tier) |>
  mutate(tier = factor(tier, levels = c("High", "Medium", "Low", "Uncertain")),
         pct = n / sum(n)) |>
  ggplot(aes(tier, n, fill = tier)) +
  geom_col() +
  geom_text(aes(label = sprintf("%s\n(%.1f%%)", comma(n), 100 * pct)), vjust = -0.2) +
  scale_fill_manual(values = c(High = "#00A087", Medium = "#7E6148",
                                Low = "#E64B35", Uncertain = "grey70"), guide = "none") +
  labs(title = "OCS tier distribution", x = NULL, y = "Count") +
  theme_minimal(base_size = 13)
```

# 5. Expression Distributions

```{r expr-dist, fig.height=10, fig.width=12, echo=FALSE}
read_cols <- grep("_reads$", names(data$expression), value = TRUE)
pN_cols   <- grep("_pN$", names(data$expression), value = TRUE)

expr_agg <- data$expression |>
  reframe(
    max_reads = do.call(pmax, c(pick(all_of(read_cols)), na.rm = TRUE)),
    max_pN    = do.call(pmax, c(pick(all_of(pN_cols)), na.rm = TRUE)),
    n_samples = rowSums(pick(all_of(read_cols)) > 0, na.rm = TRUE)
  )

# Reads histogram
p1 <- ggplot(expr_agg, aes(max_reads)) +
  geom_histogram(bins = 100, fill = "#4682B4", alpha = 0.8) +
  geom_vline(xintercept = cfg$filtering$prelim_reads_min, color = "#E64B35",
             linewidth = 1, linetype = "dashed") +
  scale_x_log10(labels = comma) +
  labs(title = "Max reads per ORF (signal presence)",
       subtitle = sprintf("Red line: reads > %d", cfg$filtering$prelim_reads_min),
       x = "max reads across samples (log10)", y = "Count") +
  theme_minimal(base_size = 12)

# pN histogram
p2 <- ggplot(expr_agg, aes(max_pN)) +
  geom_histogram(bins = 100, fill = "#7E6148", alpha = 0.8) +
  geom_vline(xintercept = cfg$filtering$prelim_pN_min, color = "#E64B35",
             linewidth = 1, linetype = "dashed") +
  labs(title = "Max pN per ORF (frame periodicity)",
       subtitle = sprintf("Red line: pN > %.1f  — NOT a P-site fraction proxy (r=-0.20 vs p_site_pct)",
         cfg$filtering$prelim_pN_min),
       x = "max pN across samples", y = "Count") +
  theme_minimal(base_size = 12)

# Sample coverage
p3 <- ggplot(expr_agg, aes(n_samples)) +
  geom_histogram(binwidth = 1, fill = "#00A087", alpha = 0.8) +
  labs(title = "Samples with reads per ORF",
       subtitle = "Heterogeneous datasets — any sample with signal is informative",
       x = "N samples with reads > 0", y = "Count") +
  theme_minimal(base_size = 12)

print(p1); print(p2); print(p3)
```

# 6. Preliminary Filter Comparison

```{r filter-matrix, echo=FALSE}
filter_matrix <- build_filter_matrix(df, filter_pattern = "^prelim_pass_")

filter_matrix |>
  select(filter_label, n_retained, pct_total, n_OCS_High, n_OCS_Medium,
         n_nonCDS, n_intergenic, n_multi_tool) |>
  knitr::kable(format = "html", digits = 1,
    col.names = c("Strategy", "Retained", "% Total", "High OCS",
                  "Medium OCS", "Non-CDS", "Intergenic", "≥2 tools"),
    caption = "Preliminary filter strategy comparison")

# Visual
filter_matrix |>
  mutate(filter_label = factor(filter_label, levels = filter_label[order(n_retained)])) |>
  ggplot(aes(n_retained, n_OCS_High)) +
  geom_point(aes(size = n_nonCDS, color = pct_total), alpha = 0.8) +
  geom_text(aes(label = filter_label), size = 3, hjust = -0.1, vjust = 0.5) +
  scale_x_log10(labels = comma) +
  scale_color_viridis_c() +
  labs(title = "Preliminary filters: retention vs OCS High overlap",
       x = "ORFs retained (log10)", y = "High OCS retained",
       size = "Non-CDS", color = "% Total") +
  theme_minimal(base_size = 12)
```

# 7. Orthogonality: Expression Filters × OCS

```{r orthog, echo=FALSE}
# Filter status for all ORFs
ct <- df |>
  mutate(
    filter_status = case_when(
      prelim_pass_combined_3 ~ "reads+pN (≥3 samples)",
      prelim_pass_combined_2 ~ "reads+pN (≥2 samples)",
      prelim_pass_combined ~ "reads+pN (≥1 sample)",
      prelim_pass_reads ~ "reads only (≥1)",
      TRUE ~ "Not passed"
    ),
    ocs_group = case_when(
      tier == "High"               ~ "OCS High",
      tier %in% c("High","Medium") ~ "OCS Medium+",
      TRUE                          ~ "OCS Low/Uncertain"
    )
  ) |>
  count(filter_status, ocs_group) |>
  tidyr::pivot_wider(names_from = ocs_group, values_from = n, values_fill = 0)

knitr::kable(ct, format = "html",
  caption = "All ORFs: expression filter × OCS tier cross-tabulation")

# Tool agreement breakdown for non-CDS
ct2 <- df |>
  filter(orf_biotype != "CDS") |>
  mutate(pass = ifelse(prelim_pass_combined, "Pass prelim filter", "Fail prelim filter")) |>
  count(pass, n_detecting) |>
  tidyr::pivot_wider(names_from = n_detecting, values_from = n, values_fill = 0,
                     names_prefix = "Ntools=")

knitr::kable(ct2, format = "html",
  caption = "Non-CDS ORFs: prelim filter × number of detecting tools")
```

# 8. Correlation Between Scoring Dimensions

```{r cor-heatmap, fig.height=6, fig.width=9, echo=FALSE}
cor_data <- df |>
  filter(orf_biotype != "CDS") |>
  select(pN, total_reads_expr, max_pN, n_samples,
         s_translation, s_agreement, s_coverage, s_periodicity,
         n_detecting, n_pass_combined) |>
  rename(
    `pN (metadata)` = pN,    `Total reads` = total_reads_expr,
    `Max pN` = max_pN,       `N samples` = n_samples,
    S_translation = s_translation, S_agreement = s_agreement,
    S_coverage = s_coverage, S_periodicity = s_periodicity,
    `N tools` = n_detecting, `N pass filter` = n_pass_combined
  )

cor_matrix <- cor(cor_data, use = "pairwise.complete.obs")

cor_matrix |>
  as.data.frame() |>
  mutate(var1 = rownames(cor_matrix)) |>
  select(var1, everything()) |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  ggplot(aes(var1, var2, fill = r)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  scale_fill_gradient2(low = "#3B4992", mid = "white", high = "#E64B35",
                       limits = c(-1, 1)) +
  labs(title = "Correlation: scoring dimensions (non-CDS ORFs incl. intergenic)",
       subtitle = "Expression metrics (reads, pN) are largely orthogonal to OCS sub-scores",
       x = NULL, y = NULL, fill = "r") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 9. Stage 1 Output: ORFs for P-site Computation

**Filter applied:** Non-CDS (incl. intergenic) + reads > `r cfg$filtering$prelim_reads_min` + pN > `r cfg$filtering$prelim_pN_min` (≥`r cfg$filtering$prelim_cross_sample` sample)

```{r prelim-output, echo=FALSE}
prelim_orfs <- df |>
  filter(orf_biotype != "CDS",
         prelim_pass_combined)

message(sprintf("ORFs passing Stage 1 preliminary filter: %s", comma(nrow(prelim_orfs))))

prelim_orfs |>
  count(orf_biotype, sort = TRUE) |>
  mutate(pct = 100 * n / sum(n)) |>
  knitr::kable(format = "html", digits = 1,
    col.names = c("Biotype", "Count", "%"),
    caption = "ORFs per biotype passing preliminary filter")

# Write BED12 for compute_psite_purity.py input
# We need the full BED12 from the unified BED filtered to prelim ORFs
unified_bed <- cfg$input$unified_bed
if (file.exists(unified_bed)) {
  message("Filtering unified BED to preliminary ORFs ...")
  prelim_ids <- unique(prelim_orfs$orf_id)

  # Read BED line by line, keep only prelim ORFs
  out_bed <- file.path(cfg$input$output_dir, "prelim_orfs_for_psite.bed")
  con_in  <- if (grepl("\\.gz$", unified_bed)) gzfile(unified_bed, "r") else file(unified_bed, "r")
  con_out <- file(out_bed, "w")
  n_written <- 0
  while (length(line <- readLines(con_in, n = 1)) > 0) {
    # BED12 col4 is the ORF ID
    parts <- strsplit(line, "\t")[[1]]
    if (length(parts) >= 4 && parts[4] %in% prelim_ids) {
      writeLines(line, con_out)
      n_written <- n_written + 1
    }
  }
  close(con_in); close(con_out)
  message(sprintf("Wrote %s ORFs to %s", comma(n_written), out_bed))
} else {
  message(sprintf("WARNING: unified BED not found at %s — cannot write BED for P-site computation", unified_bed))
}

# Also write a simple TSV with key columns for downstream use
out_tsv <- file.path(cfg$input$output_dir, "prelim_orfs.tsv")
prelim_orfs |>
  select(orf_id, chrom, start, end, strand, orf_biotype,
         total_reads_expr, max_pN, tier, n_detecting,
         length_aa, is_classified) |>
  mutate(across(where(is.numeric), ~ signif(., 5))) |>
  fwrite(out_tsv, sep = "\t")
message(sprintf("Wrote %s ORFs → %s", comma(nrow(prelim_orfs)), out_tsv))
```

# 10. ORF Length Distribution (Pre-Prelim Filter)

```{r length-dist, fig.height=8, fig.width=12, echo=FALSE}
# All non-CDS ORFs (including intergenic)
p1 <- df |>
  filter(orf_biotype != "CDS") |>
  ggplot(aes(length_aa, fill = orf_biotype)) +
  geom_histogram(bins = 80, alpha = 0.7) +
  geom_vline(xintercept = 50, color = "#E64B35", linewidth = 1, linetype = "dashed") +
  scale_fill_manual(values = biotype_colors, guide = "none") +
  scale_x_log10(labels = comma) +
  labs(title = "ORF length distribution — all non-CDS (pre-filter)",
       subtitle = sprintf("Red line: 50aa cutoff. %s ORFs ≤ 50aa",
         comma(sum(df$length_aa <= 50 & df$orf_biotype != "CDS", na.rm = TRUE))),
       x = "ORF length (amino acids, log10)", y = "Count") +
  theme_minimal(base_size = 12)

# Prelim-filtered only
p2 <- prelim_orfs |>
  ggplot(aes(length_aa, fill = orf_biotype)) +
  geom_histogram(bins = 50, alpha = 0.8) +
  facet_wrap(~ orf_biotype, scales = "free_y", ncol = 4) +
  scale_fill_manual(values = biotype_colors, guide = "none") +
  geom_vline(xintercept = 50, color = "#E64B35", linewidth = 0.5, linetype = "dashed") +
  labs(title = "ORF length distribution — prelim-filtered",
       subtitle = sprintf("reads>%d + pN>%.1f (≥%d sample). %s ORFs",
         cfg$filtering$prelim_reads_min,
         cfg$filtering$prelim_pN_min,
         cfg$filtering$prelim_cross_sample,
         comma(nrow(prelim_orfs))),
       x = "ORF length (amino acids)", y = "Count") +
  theme_minimal(base_size = 12)

# Pre vs post prelim comparison by biotype
pre_len <- df |>
  filter(orf_biotype != "CDS") |>
  select(orf_biotype, length_aa) |>
  mutate(stage = "Pre-prelim")
post_len <- prelim_orfs |>
  select(orf_biotype, length_aa) |>
  mutate(stage = "Post-prelim")
biotype_len <- dplyr::bind_rows(pre_len, post_len)

p3 <- biotype_len |>
  ggplot(aes(stage, length_aa, fill = stage)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  facet_wrap(~ orf_biotype, ncol = 4) +
  scale_fill_manual(values = c("Pre-prelim" = "grey70", "Post-prelim" = "#00A087"),
                    guide = "none") +
  labs(title = "Length distribution: pre vs post prelim filter by biotype",
       x = NULL, y = "ORF length (amino acids)") +
  theme_minimal(base_size = 12)

print(p1)
print(p2)
print(p3)
```

# 11. Summary: Stage 1 Filtering Cascade

```{r summary, echo=FALSE}
data.frame(
  Stage = c("Total unified ORFs", "With GENCODE classification",
            "Intergenic (unclassified)", "Non-CDS (incl. intergenic)",
            sprintf("reads>%d (≥%d sample)", cfg$filtering$prelim_reads_min, cfg$filtering$prelim_cross_sample),
            sprintf("reads>%d + pN>%.1f (≥%d sample)", cfg$filtering$prelim_reads_min, cfg$filtering$prelim_pN_min, cfg$filtering$prelim_cross_sample),
            "Prelim + ≤50aa"),
  Count = c(
    nrow(data$metadata),
    sum(df$is_classified),
    sum(df$orf_biotype == "intergenic"),
    sum(df$orf_biotype != "CDS", na.rm = TRUE),
    sum(df$prelim_pass_reads, na.rm = TRUE),
    sum(df$prelim_pass_combined, na.rm = TRUE),
    sum(df$prelim_pass_combined & df$length_aa <= 50, na.rm = TRUE)
  )
) |>
  knitr::kable(format = "html",
    caption = sprintf("Stage 1 filtering cascade — %s", cfg$project$label))
```

## Next Step

**Run `compute_psite_purity.py`** on the preliminary ORF set:

```bash
python3 scripts/post_analysis/compute_psite_purity.py \
  --bed {output_dir}/prelim_orfs_for_psite.bed \
  --riboseqc-dir {riboseqc_dir} \
  --output {output_dir}/psite_purity.tsv \
  --workers 8
```

This computes **exact** `p_site_GSE`, `p_site_pct`, and `p_site_pos` by
backtracking through the RiboseQC bedgraph files — no proxies. Running on
`r comma(nrow(prelim_orfs))` ORFs instead of `r comma(nrow(df))` ORFs.

Then run `02_psite_filtering.qmd` for Stage 2 final ORF selection.
