---
title: "ORF P-site Purity Filtering — `r params$project$label`"
author: "Ribo-seq Pipeline"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
params:
  config: "config_template.yaml"
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(data.table)
  library(scales)
  library(yaml)
  library(R6)
})

if (file.exists("_functions.R")) {
  source("_functions.R")
} else if (file.exists("../../scripts/post_analysis/_functions.R")) {
  source("../../scripts/post_analysis/_functions.R")
} else {
  stop("Cannot find _functions.R")
}

cfg  <- load_config(params$config)
data <- load_pipeline_data(cfg)

# ── Load prelim ORFs from Stage 1 ──
prelim_file <- file.path(cfg$input$output_dir, "prelim_orfs.tsv")
if (file.exists(prelim_file)) {
  prelim_orfs <- .read_tsv(prelim_file)
  message(sprintf("Loaded %s prelim ORFs from Stage 1", scales::comma(nrow(prelim_orfs))))
} else {
  stop("prelim_orfs.tsv not found — run 01_prelim_analysis.qmd first")
}

# ── Apply real P-site filters ──
psite_file <- file.path(cfg$input$output_dir, "psite_purity.tsv")
if (!file.exists(psite_file)) {
  stop(sprintf(
    "psite_purity.tsv not found at %s — run:\n  python3 compute_psite_purity.py --bed %s --riboseqc-dir %s --output %s --workers 8",
    psite_file,
    file.path(cfg$input$output_dir, "prelim_orfs_for_psite.bed"),
    cfg$input$riboseqc_dir,
    psite_file
  ))
}

# Merge: start from merged data, filter to prelim ORFs, apply P-site filters
df <- data$merged |>
  semi_join(prelim_orfs |> select(orf_id), by = "orf_id")

df <- apply_psite_filters(df, cfg, psite_file)

# ── Expression filter context (for cross-comparison) ──
# Filter expression to match the working set
expr_filtered <- data$expression |> semi_join(df |> select(orf_id), by = "orf_id")
df <- apply_prelim_filters(df, cfg, expr_filtered)

message(sprintf("Working set: %s ORFs (prelim-filtered)", scales::comma(nrow(df))))
```

# 1. P-site Purity Data Overview

The P-site purity metrics are computed by `compute_psite_purity.py`,
which **backtracks through actual RiboseQC bedgraph files** — intersecting
each ORF's genomic coordinates with `_P_sites_{plus,minus}.bedgraph` and
`_coverage_{plus,minus}.bedgraph` intervals. No proxies, no approximations.

```{r psite-overview, echo=FALSE}
psite_data <- data$purity
if (!is.null(psite_data)) {
  psite_cols <- grep("_p_site_GSE$", names(psite_data), value = TRUE)
  pct_cols   <- grep("_p_site_pct$", names(psite_data), value = TRUE)
  pos_cols   <- grep("_p_site_pos$", names(psite_data), value = TRUE)
  n_samples  <- length(psite_cols)

  data.frame(
    Metric = c("ORFs with P-site data", "Samples processed", "Columns per sample",
               "Median p_site_GSE", "Median p_site_pct", "Median p_site_pos"),
    Value = c(
      scales::comma(nrow(psite_data)),
      as.character(n_samples),
      "p_site_GSE, p_site_pct, p_site_pos, reads_GSE, not_p_site_GSE",
      sprintf("%.1f", median(rowSums(psite_data[, psite_cols, drop = FALSE], na.rm = TRUE))),
      sprintf("%.3f", median(rowMeans(psite_data[, pct_cols, drop = FALSE], na.rm = TRUE), na.rm = TRUE)),
      sprintf("%.1f", median(rowMeans(psite_data[, pos_cols, drop = FALSE], na.rm = TRUE), na.rm = TRUE))
    )
  ) |>
    knitr::kable(format = "html", caption = "P-site purity data summary")
} else {
  cat("**ERROR: P-site purity data not loaded.** Check psite_purity.tsv.\n")
}
```

# 2. P-site Metric Distributions

```{r psite-dist, fig.height=12, fig.width=12, echo=FALSE}
if (!is.null(psite_data)) {
  psite_cols <- grep("_p_site_GSE$", names(psite_data), value = TRUE)
  pct_cols   <- grep("_p_site_pct$", names(psite_data), value = TRUE)
  pos_cols   <- grep("_p_site_pos$", names(psite_data), value = TRUE)

  # Max per ORF across samples
  psite_agg <- psite_data |>
    reframe(
      max_gse = do.call(pmax, c(pick(all_of(psite_cols)), na.rm = TRUE)),
      max_pct = do.call(pmax, c(pick(all_of(pct_cols)), na.rm = TRUE)),
      max_pos = do.call(pmax, c(pick(all_of(pos_cols)), na.rm = TRUE)),
      n_psite_samples = rowSums(pick(all_of(psite_cols)) > 0, na.rm = TRUE)
    )

  filt <- cfg$filtering

  p1 <- ggplot(psite_agg, aes(max_gse)) +
    geom_histogram(bins = 100, fill = "#4682B4", alpha = 0.8) +
    geom_vline(xintercept = filt$p_site_gse_min, color = "#E64B35",
               linewidth = 1, linetype = "dashed") +
    scale_x_log10(labels = comma) +
    labs(title = "Max p_site_GSE per ORF (P-site reads)",
         subtitle = sprintf("Red line: GSE > %d | Pass: %s ORFs (%.1f%%)",
           filt$p_site_gse_min,
           comma(sum(psite_agg$max_gse > filt$p_site_gse_min)),
           100 * mean(psite_agg$max_gse > filt$p_site_gse_min)),
         x = "max p_site_GSE across samples (log10)", y = "Count") +
    theme_minimal(base_size = 12)

  p2 <- ggplot(psite_agg, aes(max_pct)) +
    geom_histogram(bins = 100, fill = "#00A087", alpha = 0.8) +
    geom_vline(xintercept = filt$p_site_pct_min, color = "#E64B35",
               linewidth = 1, linetype = "dashed") +
    labs(title = "Max p_site_pct per ORF (P-site fraction)",
         subtitle = sprintf("Red line: pct > %.1f | Pass: %s ORFs (%.1f%%)",
           filt$p_site_pct_min,
           comma(sum(psite_agg$max_pct > filt$p_site_pct_min)),
           100 * mean(psite_agg$max_pct > filt$p_site_pct_min)),
         x = "max p_site_pct across samples", y = "Count") +
    theme_minimal(base_size = 12)

  p3 <- ggplot(psite_agg, aes(max_pos)) +
    geom_histogram(bins = 100, fill = "#7E6148", alpha = 0.8) +
    geom_vline(xintercept = filt$p_site_pos_min, color = "#E64B35",
               linewidth = 1, linetype = "dashed") +
    labs(title = "Max p_site_pos per ORF (P-site position offset)",
         subtitle = sprintf("Red line: pos > %d | Pass: %s ORFs (%.1f%%)",
           filt$p_site_pos_min,
           comma(sum(psite_agg$max_pos > filt$p_site_pos_min)),
           100 * mean(psite_agg$max_pos > filt$p_site_pos_min)),
         x = "max p_site_pos across samples", y = "Count") +
    theme_minimal(base_size = 12)

  p4 <- ggplot(psite_agg, aes(n_psite_samples)) +
    geom_histogram(binwidth = 1, fill = "#F39B7F", alpha = 0.8) +
    geom_vline(xintercept = filt$final_cross_sample, color = "#E64B35",
               linewidth = 1, linetype = "dashed") +
    labs(title = "Samples with P-site reads per ORF",
         subtitle = sprintf("Red line: ≥%d samples required",
           filt$final_cross_sample),
         x = "N samples with p_site_GSE > 0", y = "Count") +
    theme_minimal(base_size = 12)

  print(p1); print(p2); print(p3); print(p4)
}
```

# 3. Proxy Validation: Expression vs Real P-site

This section confirms the proxy validation from earlier analysis:
`reads` is an excellent proxy for `p_site_GSE` (r ≈ 0.99), while
`pN` is NOT a proxy for `p_site_pct` (r ≈ -0.20).

```{r proxy-validation, fig.height=6, fig.width=12, echo=FALSE}
if (!is.null(psite_data)) {
  # Merge P-site data with expression data
  comp <- df |>
    filter(!is.na(total_reads_expr), !is.na(psite_pass_gse)) |>
    select(orf_id, total_reads_expr, max_pN, total_psites_real, mean_p_site_pct,
           n_pass_gse, n_pass_pct, n_pass_triple)

  if (nrow(comp) > 0) {
    comp <- comp |> filter(!is.na(total_psites_real), total_reads_expr > 0)

    # Reads vs p_site_GSE
    p1 <- ggplot(comp |> filter(total_psites_real > 0),
           aes(total_reads_expr, total_psites_real)) +
      geom_point(alpha = 0.1, size = 0.5, color = "#4682B4") +
      geom_smooth(method = "lm", color = "#E64B35", se = FALSE) +
      scale_x_log10(labels = comma) +
      scale_y_log10(labels = comma) +
      labs(title = sprintf("Expression reads vs Real P-site GSE (r=%.3f)",
             cor(log10(comp$total_reads_expr + 1), log10(comp$total_psites_real + 1))),
           x = "Total reads (expression summary, log10)",
           y = "Total p_site_GSE (bedgraph, log10)") +
      theme_minimal(base_size = 12)

    # pN vs p_site_pct
    p2 <- ggplot(comp, aes(max_pN, mean_p_site_pct)) +
      geom_point(alpha = 0.1, size = 0.5, color = "#00A087") +
      geom_smooth(method = "lm", color = "#E64B35", se = FALSE) +
      labs(title = sprintf("pN (periodicity) vs Real P-site fraction (r=%.3f)",
             cor(comp$max_pN, comp$mean_p_site_pct, use = "pairwise.complete.obs")),
           subtitle = "pN measures frame periodicity, NOT P-site fraction — pN is NOT a valid proxy for p_site_pct",
           x = "max pN (expression summary)", y = "Mean p_site_pct (bedgraph)") +
      theme_minimal(base_size = 12)

    print(p1); print(p2)
  }
}
```

# 4. Real P-site Filter Comparison

```{r psite-filter-matrix, echo=FALSE}
# Combine prelim and psite filter columns
filter_matrix <- build_filter_matrix(df, filter_pattern = "^(prelim_pass_|psite_pass_)")

filter_matrix |>
  select(filter_label, n_retained, pct_total, n_OCS_High, n_OCS_Medium,
         n_nonCDS, n_intergenic, n_multi_tool) |>
  knitr::kable(format = "html", digits = 1,
    col.names = c("Strategy", "Retained", "% Total", "High OCS",
                  "Medium OCS", "Non-CDS", "Intergenic", "≥2 tools"),
    caption = "Prelim + P-site filter strategy comparison")

filter_matrix |>
  mutate(filter_label = factor(filter_label, levels = filter_label[order(n_retained)])) |>
  ggplot(aes(n_retained, n_OCS_High)) +
  geom_point(aes(size = n_nonCDS, color = pct_total), alpha = 0.8) +
  geom_text(aes(label = filter_label), size = 3, hjust = -0.1, vjust = 0.5) +
  scale_x_log10(labels = comma) +
  scale_color_viridis_c() +
  labs(title = "Filter strategies (prelim + P-site): retention vs OCS High",
       x = "ORFs retained (log10)", y = "High OCS retained",
       size = "Non-CDS", color = "% Total") +
  theme_minimal(base_size = 12)
```

# 5. Final ORF Selection

**Filter:** Non-CDS (incl. intergenic) + real P-site GSE > `r cfg$filtering$p_site_gse_min`
+ p_site_pct > `r cfg$filtering$p_site_pct_min`
(≥`r cfg$filtering$final_cross_sample` samples) + length ≤ 50 amino acids.

> **Note:** `p_site_pos` is not computed by the fast bedtools-based P-site computation.
> The position filter is excluded — it is the least discriminating criterion anyway
> (>97% pass rate in prior validation).

```{r final-selection, echo=FALSE}
# Apply final filter (GSE + pct; pos not computed by fast bedtools mode)
final_orfs <- df |>
  filter(orf_biotype != "CDS",
         psite_pass_gse,
         psite_pass_pct,
         length_aa <= 50)

message(sprintf("Final ggRibo candidates: %s ORFs", scales::comma(nrow(final_orfs))))

final_orfs |>
  count(orf_biotype, sort = TRUE) |>
  mutate(pct = 100 * n / sum(n)) |>
  knitr::kable(format = "html", digits = 1,
    col.names = c("Biotype", "Count", "%"),
    caption = "Per-biotype ORF counts passing all filters")

# Write final ORF list for ggRibo
out_tsv <- file.path(cfg$input$output_dir, "final_orfs_for_ggribo.tsv")
final_orfs |>
  select(orf_id, chrom, start, end, strand, orf_biotype, gencode_biotype,
         total_reads_expr, max_pN, total_psites_real, mean_p_site_pct,
         tier, ocs, n_detecting, length_aa, is_classified) |>
  mutate(across(where(is.numeric), ~ signif(., 5))) |>
  fwrite(out_tsv, sep = "\t")

message(sprintf("Wrote %s ORFs → %s", scales::comma(nrow(final_orfs)), out_tsv))
```

# 6. Final Candidate Summary

```{r final-summary, fig.height=8, fig.width=12, echo=FALSE}
# OCS boxplot by biotype
if (nrow(final_orfs) > 0 && "ocs" %in% names(final_orfs) && "orf_biotype" %in% names(final_orfs)) {
  bt_cols <- c(intORF = "#E64B35", dORF = "#00A087", doORF = "#7E6148",
               uORF = "#4682B4", uoORF = "#F39B7F", lncRNA = "#3C5488",
               intergenic = "#FFDC91")
  present_bt <- intersect(unique(as.character(final_orfs$orf_biotype)), names(bt_cols))
  boxplot(ocs ~ orf_biotype, data = final_orfs,
          col = bt_cols[present_bt],
          main = "OCS Distribution by ORF Biotype (Final Candidates)",
          xlab = NULL, ylab = "ORF Confidence Score (OCS)",
          las = 2, cex.axis = 0.9)
  abline(h = c(0.2, 0.5, 0.8), lty = 2, col = "grey60")
}

# Tier composition
tt <- as.data.frame(table(final_orfs$orf_biotype, final_orfs$tier))
names(tt) <- c("biotype", "tier", "n")
tt$tier <- factor(tt$tier, levels = c("High", "Medium", "Low", "Uncertain"))

ggplot(tt, aes(biotype, n, fill = tier)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = c(High = "#00A087", Medium = "#7E6148",
                                Low = "#E64B35", Uncertain = "grey70"),
                    name = "OCS Tier") +
  labs(title = "OCS tier composition by ORF biotype",
       subtitle = sprintf("Final candidates: %s ORFs", scales::comma(nrow(final_orfs))),
       x = NULL, y = "Proportion") +
  theme_minimal(base_size = 12)

# Multi-tool support
final_orfs |>
  count(n_detecting) |>
  mutate(pct = 100 * n / sum(n)) |>
  knitr::kable(format = "html", digits = 1,
    col.names = c("N detecting tools", "Count", "%"),
    caption = "Multi-tool support for final candidates")
```

# 7. Final Filtering Cascade

```{r cascade, echo=FALSE}
data.frame(
  Stage = c(
    "Total unified ORFs",
    "Non-CDS (incl. intergenic)",
    "Stage 1: Non-CDS + reads+pN prelim",
    sprintf("Stage 2: + p_site_GSE > %d (≥%d samples)", cfg$filtering$p_site_gse_min, cfg$filtering$final_cross_sample),
    sprintf("Stage 2: + p_site_pct > %.1f (≥%d samples)", cfg$filtering$p_site_pct_min, cfg$filtering$final_cross_sample),
    "Final: + length ≤ 50aa"
  ),
  Count = c(
    nrow(data$merged),
    sum(data$merged$orf_biotype != "CDS", na.rm = TRUE),
    sum(df$prelim_pass_combined, na.rm = TRUE),
    sum(df$psite_pass_gse, na.rm = TRUE),
    sum(df$psite_pass_gse & df$psite_pass_pct, na.rm = TRUE),
    nrow(final_orfs)
  )
) |>
  knitr::kable(format = "html",
    caption = sprintf("Two-stage filtering cascade — %s", cfg$project$label))
```

## Next Step

Run `03_generate_ggribo.qmd` to generate ggRibo coverage plots for
the `r scales::comma(nrow(final_orfs))` final candidates,
with top `r cfg$ggribo$top_n_samples` samples per ORF sorted by
`r cfg$ggribo$sort_metric`.
